Embedded Domain Specific Languages inside Functional Languages

### 2014-03-06 paper.tex
Hacking paper

\pagebreak
Haskell is a concise and direct language.
Structures in Haskell are described using types, constructed
and deconstructed, but never updated. The entire language
functions by chaining together these structural processors,
which ultimately take input, and produce output. Side-effects
are described using a |do|-notation, for example:

@@

In this example a {\em value\/} called |main| uses the |do|-notation to describe
an interaction with a user. Actually, the |do|-notation captures
this as a structure called a Monad; purity is not compromised. For more details
about how |do|-notation and Monads can provide an effectful interface
inside a pure language like Haskell, see~\cite{SPJ:93:IFP}. For the
purposes of Kansas Lava, |do|-notation is a way of providing syntax
and structure that looks like interaction.

@@

@@

Another observation to make regarding EDSLs is that
there are two flavors:
%% I don't like the use of ``First'' and ``Second'' in these bullet points.
\begin{itemize}
\item First, {\em DSLs that use a shallow embedding\/}, where values are computed with directly.
The result of a computation in a shallow DSL is a value.
\item Second, {\em DSLs that use a deep embedding\/} build an abstract syntax tree~\cite{Elliott:03:CompileDSEL-JFP}.
The result of a computation inside a deep DSL
is a structure, not a value, and this structure can be used to compute a value,
or be cross-compiled before being evaluated.
\end{itemize}
This is an important distinction. Historically, EDSLs have been shallow,
and as such are simply a ways of structuring an API to a library. Deep EDSLs
have the ability to {\em stage\/} code, that is executing a program
can generate another program, much like the well-known YACC DSL,
but at the cost of significantly restricting what forms of the DSL can
generate valid output from.
%% That sentence is a bit garbled.

The unifying theme is that embedded DSLs can
be pragmatic, productive and useful. We present
these to allow us to make a technical argument that adding rewrite systems to
DSLs will improve their usefulness by removing
some of the inherent and outstanding shortcomings.

@@

A deeply embeeded DSL can be ...

\subsection{How to extract a deep embedding}

The basis of

@@

>>>>

Haskell is a concise and direct language.
Structures in Haskell are described using types, constructed
and deconstructed, but never updated.

(( ABOUT DATA Construct ))

data Maybe where
  Nothing ::      Maybe a
  Just    :: a -> Maybe a

The entire language
functions by chaining together these structural processors,
which ultimately take input, and produce output. Side-effects
are described using a |do|-notation, for example:

@@

In this example a {\em value\/} called |main| uses the |do|-notation to describe
an interaction with a user. Actually, the |do|-notation captures
this as a structure called a Monad; purity is not compromised. For more details
about how |do|-notation and Monads can provide an effectful interface
inside a pure language like Haskell, see~\cite{SPJ:93:IFP}. For the
purposes of this article, |do|-notation is a way of providing syntax
and structure that looks like interaction.

@@

Compositionally
Combining elements into more complex ones should be easy and natural.
Abstraction
The user shouldn’t have to know (or be allowed to exploit) the underlying implementation of your types.

@@

There are two flavors of EDSLs.
The most common flavor are
{\em DSLs that use a shallow embedding\/}, where values are computed with directly.
The result of a computation in a shallow DSL is a value.
But there is another class of DSLs, 
specifically {\em DSLs that use a deep embedding\/} build an abstract syntax tree~\cite{Elliott:03:CompileDSEL-JFP}.
The result of a computation inside a deep DSL
is a structure, not a value, and this structure can be used to compute a value,
or be cross-compiled before being evaluated.

This is an important distinction. Historically, EDSLs have been shallow,
and as such are simply a ways of structuring an API to a library. Deep EDSLs
have the ability to {\em stage\/} code, that is executing a program
can generate another program, much like the well-known YACC DSL,
but at the cost of significantly restricting what forms of the DSL can
generate valid output from.
%% That sentence is a bit garbled.

The unifying theme is that embedded DSLs can
be pragmatic, productive and useful.

@@

A deeply embeeded DSL is a DSL than can have ???.
To make this concreate, consider a DSL for arithmetic,
with addition, subtraction, multiplcation, and constants.
We want to be able to {\bf run\/} this DSL. For a shallow
embedding this is trival - we just use the built in 
arithmetic. A deep embedding is where things get
interesting. Consider a data-type for our arithmetic.

\begin{Code}
data E where
 Lit :: Integer  -> E
 Add :: E -> E -> E
 Sub :: E -> E -> E
 Mul :: E -> E -> E
 deriving Eq
\end{Code}

Now we overload the arithemtic to use this E datatype;
in Haskell Num is our overloading for integeral arithmetic.

\begin{Code}
instance Num E where
  e1 + e2 = Add e1 e2
  ...

Now, if we build expressions of type E, we can
observe the structure of the computation.
\begin{SemiCode}
GHCi> 1 + 2 * 3
Mul (Add (Lit 1) (Lit 2)) (Lit 3)
\end{SemiCode}

With deep embeddings, it is common to also
write a run function, that computes the structure
of a catured computation, given a simple value.
\begin{Code}
run :: E -> Int
run (...)
\end{Code}

About running the code.

\subsection{How to extract a deep embedding from a Function}

Expressing function calls in terms of constructors,  and
building expression trees is a gimic. THere are entire
systems build on the idea of building structures 
in principaled ways, for example JHDL~\cite{..}, but
we can do much more. With careful construction,
we can capture function defintions,
and other synatically structuruals,
directly from a deep embedding. It is at this
point, the idea of capturing code,
the using the captured code to execute code
on a different target becomes possible.

f :: E -> E
f x = x + 1

Here we have a function that acts over our new type
\verb|E|, and returns a new \verb|E|. How can we capture
this function.
The trick is to invent a unique E, and pass it as the argument
to \verb|f|.

\begin{Code}
data E where
  Lit :: Integer -> E
  ...
  Var :: String -> E    -- new constructor
\end{Code}

\begin{Code}
-- Just running the function
> f 4
Add (Lit 4) (Lit 1)
-- reifing the function, using our unique E ``Var''.
> f (Var "x")
Add (Var "x") (Lit 1)   -- reified verision of the function
\end{Code}

This is remarkabile. We're run a function with 
a dummy argument (called the protypical argument)
and extracted the body of the function.

The obvious questions are does this trick used here scale,
and what are the limitations? The trick has
been used on may large examples, including
generating large telemerty circuits, 
and (( .. feldspar )). It is considered a
stardard pattern in the functional programming
community, analogoues to the OO patterns
enumerated in the gang of four book~\cite{..}.
The trick is made all the more appealing
because we can use operator overloading.
But the trick does has serious limitations.

The basis of a Deep EDSL is one of constructiveness.
The aim is to construct a structure, then
use that structure that was previously implicit.
Figure sumurizes the state of the art with
deeply embeeded reification

\begin{verbatim}
Expressions             Natural Fit
Functions               The trick
local bindings          Yes, using observable sharing
internal functions      Yes, using extension of OS
do-statements           Yes, using monadic reification
control flow            Not directly, can simulate
\end{verbatim}

However, the trick breaks down on control flow,
which is not constructive, but instead requires
answers.

@@

Lava programs are written as equations of recusive bindings.
